---
title: "Linear mixed models to RDF"
subtitle: "Generating examples for the paper"
output:
  html_document:
    toc: true
    toc_float: true
    warning: false
    message: false
    df_print: paged
  html_notebook: 
    theme: readable
    toc: yes
editor_options: 
  chunk_output_type: inline
---

This is a demo instruction of how to produce a semantic model (RDF graph) from results of statistical modelling by linear mixed models in R (lme4 and nlme packages).
The tool for RDF graph generation _semLMM_ currently supports a subset of models that can be declared and fitted in these packages.

### Library semLMM
The code for genaration of RDF graphs from linear mixed models in R, is being developed as an R package _semLMM_. To be able to use it, install and attach _semLMM_ library, which is contained in the Github repository 'SemanticLMM'.

```{r}
library(semLMM)
```

#### Example datasets
Two exemplary datasets, given in the paper that describes this work, are attached to the library. *Example1* is a toy dataset of 6 observations of a variable 'y' for objects characterised by variables 'Treatment' and 'Block'.
```{r}
example1
```
*Example3* is a real-size dataset, obtained from a [plant experiment](http://cropnet.pl/plantphenodb/index.php?id=250) in which 10 characteristics
(TGW, GW_m2, GW_MS, HD, main_spike, main_stem, G_MS, TL, Wvpd, Precipitation) were observed for 611 objects, classified by 3 factors (InfraspecificName, StudyStart, Replication).
```{r}
example3
```



### Generating graphs for R lmer models
Generating RDF graphs for R models of type lmer from *lme4* package or *lmerTest* package (to include testing for effects):
```{r}
require(lmerTest)
#require(lme4)
```

#### Modelling for Example 1

1. Define a statistical model of type lmer for example1 dataset. Here, we declare a model with no intercept, single fixed term for Treatment, and single random term for Block:
```{r warning=FALSE}
mod <- lmer(y ~ 0 + Treatment + (1|Block), data = example1)
summary(mod)
anova(mod)
```

2. Build a semantic model of a statistical model:
```{r warning = FALSE}
modelFitting <- exportModelToRDF(mod)
```

3. Serialize the annotated model as RDF triples:
```{r}
modelFitting$saveTriples(graphName = "example1")
```

4. View the result: preview RDF for the model fitting process (only the main concept, without expanding the linked entities):
```{r}
cat(modelFitting$asTTL())
```

Preview the whole graph for the main object (expanding to all related entities): 
```{r}
print(modelFitting)
```


### Generating graphs for R lme models
Generating RDF graphs for R models of type lme from *nlme* package in done in the same way as above:
```{r message=FALSE, warning=FALSE}
require(nlme)
```

#### Modelling for Example 1

1. Specify the model according to lme syntax. Here, we assume the same model as above: no intercept term, single fixed term for Treatment, single random terms for Block. 
```{r}
require(nlme)
mod <- lme(y ~ 0 + Treatment, random = ~1|Block, data = example1)
```

Build a semantic model of a statistical model, and serialise RDF to a file.
```{r warning=FALSE}
modelFitting <- exportModelToRDF(mod, example1)
modelFitting$saveTriples(graphName = "example1")
```


#### Modelling for Example 3
In the exemplary models for example3 we use a subset of data (by selecting 4 initial values of the categorical variable InfraspecificName) to produce RDF graphs of a convenient (tractable) size.

##### Model with single fixed term and single random term
```{r warning=FALSE}
mod <- lme(TGW ~ InfraspecificName, random = ~1|Replication, 
           data = example3, subset = as.numeric(InfraspecificName) < 5)
summary(mod)
anova(mod)

modelFitting <- exportModelToRDF(mod, ds = example3)
modelFitting$saveTriples()
```

##### Model with two fixed terms for StudyStart and HD (regressive), and signle random terms for Replication.
```{r warning=FALSE}
mod <- lme(GW_m2 ~ 0 + StudyStart + HD, random = ~1|Replication, 
           data = example3, subset = as.numeric(InfraspecificName) < 5)
summary(mod)
anova(mod)

modelFitting <- exportModelToRDF(mod, ds = example3)
modelFitting$saveTriples()
```

##### Model with two fixed terms, interaction, regression and single random term
The full interaction (denoted in R shortcut notation as 'InfraspecificName*StudyStart') is unfolded to three terms: InfraspecificName + StudyStart + InfraspecificName:StudyStart. This is reflected in the semantic model. 

```{r warning=FALSE}
mod <- lme(GW_m2 ~ InfraspecificName*StudyStart + HD, random = ~1|Replication, 
           data = example3, subset = as.numeric(InfraspecificName) < 5)
summary(mod)
anova(mod)

modelFitting <- exportModelToRDF(mod, ds = example3)
modelFitting$saveTriples()
```

### Generating graphs manually
Generating RDF graphs for other models, whose automated processing is not supported by semLMM library, requires manual parsing of the model outputs and calling semLMM functions to create individual LMM entities and link them together. After such a model is built, saveTriples() function can be used to serialize it the RDF graph in .trig file.

Custom parsing and model creation is used to process models obtained from Genstat software (see example2 in the paper).

```{r}
# sample code, not run
modelFitting <- Process("modelFittingGenstat", processType="ModelFitting", 
                        comments = list("rdfs:comment 
                                        \"Results from software Genstat 19.1 (VSNi Ltd.)\""))
lmm <- Lmm(label="model_y1y2y3y4-0.Treatment", 
           formula = "VCOMPONENTS [FIXED=%_multiY.%_Treatment; CONSTANT=omit]; 
           random = %_multiY.%_Block + %_multiY.%_units; CONSTRAINTS=positive; 
           VSTRUCTURE [TERMS=%_multiY.%_Treatment] FACTOR=%_multiY; MODEL=ar; 
           VSTRUCTURE [TERMS=%_multiY.%_units] FACTOR=%_multiY; MODEL=diag;")

modelFitting$hasInput <- append(modelFitting$hasInput, lmm)
# {...}
```

